# Communicable, Maternal, Neonatal, and Nutritional Disease Burden (CMNN)

```{r setup, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
```


```{r dataset}
# Load necessary packages
pacman::p_load(tidyverse, knitr, here)

# Import the communicable diseases data
data_cmnn <- read_csv(here("data/burden-of-disease-cmnn.csv"))

```

```{r selected countries}
# Filter the dataset to include only the three countries chosen by your group.

# Chosen countries: Ghana (GHA), Burkina Faso (BFA), Cote d'Ivoire(CIV)

data_cmnn_mini <- data_cmnn %>% 
  filter(Code %in% c("CIV", "GHA", "BFA")) %>% 
  rename(dalys = `DALYs (Disability-Adjusted Life Years) - Communicable, maternal, neonatal, and nutritional diseases - Sex: Both - Age: Age-standardized (Rate)`)

```
Communicable, Maternal, Neonatal, and Nutritional diseases (CMNN) represent a group of health conditions that significantly contribute to the global burden of disease, particularly in low- and middle-income countries.

## Table of Estimates for CMNN Burden Over Time

```{r table}

# Here render a table for the DALY burden over time for the three countries.
# You should pivot the data wider to show each country in a separate column.

data_cmnn_wide <- data_cmnn_mini %>% 
   pivot_wider(
    names_from = Year,
    values_from = dalys  
  )

# Use kable() from the knitr package to render the table.
kable(data_cmnn_wide,
      col.names = c("Country", "Country Code", as.character(1990:2019)),
      caption = "DALYS by Country and Year")

```

## Plot Showing Trends in CMNN Burden Over Time in West Africa

Ghana, Burkino Faso and Cote d'Ivoire are three West African countries that started out with significantly different numbers of DALYs but due to continous efforts in health care have aligned over the past 20 years. 

Communicable, Maternal, Neonatal, and Nutritional diseases. These represent a group of health conditions that significantly contribute to the global burden of disease, particularly in low- and middle-income countries.

```{r}
# Here, plot a line graph showing the trend of DALY burden for the chosen countries over time. You can use ggplot2 to create this plot. Each country's line should be a different color.

ggplot(data_cmnn_mini, 
       mapping = aes(x = Year, 
                     y = dalys,
                     color = Entity)) +
  geom_line() +
  geom_point() +
  labs(title = "DALYs (Disability Adjusted Life Years) over time",
       x = "Year",
       y = "DALYs (in Persons Years)",
       color = "Countries") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5) # Bold and centered title
  )
# NOTE: Use your long 'mini' data frame for plotting, not the wide data used for the table. 
```

## Summary of CMNN Burden in West Africa

Between 1990 and 2019, Disability-Adjusted Life Years (DALYs) experienced a marked reduction across Burkina Faso, Côte d'Ivoire, and Ghana. Burkina Faso initially recorded the highest DALYs in 1990; however, by 2019, its DALYs had declined significantly, reaching levels comparable to those observed in Côte d'Ivoire. Notably, for over a decade during this period, DALYs in Burkina Faso fell below those of Côte d'Ivoire. In 1990, Côte d'Ivoire had DALYs situated between those of Ghana and Burkina Faso, yet its DALYs spiked sharply post-1990, peaking around the year 2000. Following this peak, Côte d'Ivoire experienced a reduction in DALYs, ultimately aligning with those of Burkina Faso. Ghana, on the other hand, began with substantially lower DALYs in 1990, demonstrating a gradual and steady decline from approximately 40,000 person-years in 1990 to 20,000 person-years by 2019. Collectively, these three nations have made remarkable progress in reducing DALYs over the examined timeframe.
